Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-11-11 paper/type-inference.tex
Edits.

Region type templates are FGJ types extended with fresh region
variables ($\rho$) and predicate variables ($\varphi$) to denote
unknown region annotations and region constraints, respectively. For
instance, if a variable $x$ has type $\ObjZ$ in FGJ, its region type
template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
region variable. Likewise, given the region-annotated definition of
\C{Pair} class from \S\ref{sec:fb-syntax} a region type template for a
method with FGJ type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$
is \footnote{In our exposition, we assume that classes $\C{A}$ and
$\C{B}$ are trivial subclasses of $\ObjZ$ with no fields/methods. Like
$\ObjZ$, they accept one region parameter - the allocation region of
their objects.}\footnote{We abuse arrow notation to also represent
types of methods, but unlike function types, there is no allocation
region annotation atop the arrow in a method type.}
$\inang{\rhoalloc_0,\rho_1,\rho_2,\rho_3 \,|\, \varphi_0}
\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_1,\rho_2,\rho_3} \rightarrow
\unitZ$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
variables, and $\varphi_0$ is a fresh predicate variable denoting
unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.  Region type
template of a type variable $\C{a}$ is $\C{a}@\rho$, where $\rho$ is
fresh. For a class, a template is computed in two steps. In the first
step, we templatize the types of all its fields, constructor
arguments, and arguments and return values of all its methods, along
with the type of its superclass. In the second step, we generalize all
the free region variables occuring in the templates computed in the
first step as region parameters of the class. Finally, we add a fresh
allocation region parameter ($\rhoalloc$) to the list of parameters,
and introduce a new predicate variable ($\varphi$) to denote unknown
constraints on region parameters. For example, consider the standard
FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
respectively. It can be templatized as following:
\begin{center}
\begin{codejava}
class Pair$\inang{a\extends\ObjZ, b\extends\ObjZ}$
          $\inang{\rhoalloc_0,\rho_{0-4}\,|\,\varphi_0} \extends \ObjZ\inang{\rho_4}$ {
  $a@\rho_0$ fst;
  $b@\rho_1$ snd;
  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { $...$ }
  $...$
}
\end{codejava}
\end{center}
% \footnote{In reality, the generalized region variables only come
% from the type templates of class's fields and superclass. Method
% types do not contain any free region variables.}
The template elides $\C{Pair}$'s methods, whose region type templates
contain no free variables. Among the region parameters of the class
template, $\rho_{0-4}$ are obtained by generalizing free region
variables in the types of its class's fields, constructor arguments,
and its superclass type.  The remaining parameter ($\rhoalloc_0$) is a
fresh region variable denoting the allocation region argument. Fresh
predicate variable $\varphi_0$ denotes unknown constraints over
$\rhoalloc_0$ and $\rho_{0-4}$ that need to hold for template to be a
well-formed region-annotated class definition in $\FB$.

The type template for a recursively defined class is computed in two
steps. First, all recursive occurances of the class among the types of
its fields are ignored and the class is templatized as if it is a
non-recursive class. Next, all the recursive occurances are
templatized with respect to the class template computed in the first
step, such that their region annotations are exactly same as the
region parameters of the class. For example, consider a generic
$\C{ListNode}\inang{\C{a}}$ class containing two fields: $\C{data}$ of
type $\C{a}$ and $\C{next}$ of type $\C{ListNode}\inang{\C{a}}$.
where $\C{a}$ is the type of the data stored in the linked list node.
To templatize the $\C{ListNode}\inang{a}$ class, we first ignore its
recursive occurance in the type of $\C{next}$ field, and templatize
the type $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this
type template of $\C{data}$ field, we compute the class's template as
following:
\begin{center}
$\C{ListNode}\inang{\C{a}\extends\ObjZ}
\inang{\rhoalloc_0,\rho_0\,|\,\varphi_0} \extends \ObjZ \{...\}$
\end{center}
Next, we trivially templatize the type of $\C{next}$ field in the body
of the class as $\C{ListNode}\inang{\C{a}}\inang{\rhoalloc_0,\rho_0}$.
The resulting class represents a linked list with spine in the region
$\rhoalloc_0$ and data objects in the region $\rho_0$.

% Observe that our strategy of trivially templatizing the recursive
% occurances of a class type in its definition ensures that no new
% region variables are introduced.
The templatization technique we described above for recursive class
definitions can be extended to mutually recursive definitions in a
straightforward manner, by simultaneously templatizing them. Using the
techniques outlined above, we compute region type templates for all
classes bound in the class table of the $\absof{\FB}$ program before
we proceed to elaborate expressions.

@@

Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
this elaboration for a subset of expressions in $\absof{\FB}$, whose
corresponding $\FB$ expressions have been ascribed static semantics in
Fig.~\ref{fig:fb-staticsem}. $\elabExpr$ is defined under the same
context as the expression typing judgment in
Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
retaining their meaning. The function traverses expressions in a
syntax-directed manner of a type checker, introducing fresh region
type templates for unknown region types, while generating constraints
over region and predicate variables. The precise nature of generated
constraints is explained in \S\ref{sec:fb-constraintsem}, but in
summary, they capture the relationships between the type templates of
various subexpressions and their well-formedness. Note that
$\elabExpr$ returns the region type template of the subexpression,
which is used to generate constraints for the expression. Functions
$\typeOk$ and $\subtypeOk$ (definitions not shown) used by $\elabExpr$
implement type well-formedness and subtype judgments from
Fig.~\ref{fig:fb-staticsem}, respectively.

@@

Functions $\elabMeth$ and $\elabClass$ shown in
Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
class definitions, respectively. Both functions first build a context
($\A$) containing a set ($ \rhoenv$) of region variables denoting
regions that are currently live, a map ($\aenv$) mapping type
variables to their bounds, and a constraint formula ($\phicx$)
capturing constraints over live region variables. We use predicate
variables ($\varphi$ and $\varphi_m$) to capture constraints over
variables in $\rhoenv$ denoting the fact that such constraints are yet
to be inferred.

Function $\elabMeth$ elaborates a method definition of class $B$. It
calls $\elabExpr$ with the context $\A$, its allocation context
parameter ($\rhoallocm$), and a type environment ($\env$) that
contains region type bindings for all the arguments of the method,
including the implicit $\C{this}$ argument. The region type template
returned by $\elabExpr$ for the method body is checked against its
expected type (derived from the type template of the method)
generating more constraints. The function then returns the elaborated
method definition and the set of constraints.

$\elabClass$ elaborates the definition of a class $B$. It relies on
$\elabCons$\footnote{The definition of $\elabCons$ is straightforward,
hence not shown.} and $\elabMeth$ functions to elaborate $B$'s
constructor ($k$) and method definitions ($\bar{d}$), respectively. To
the set of constraints returned by these functions, $\elabClass$ adds
constraints generated by checking the well-formedness of the type
templates of its superclass and fields, and also a new constraint
capturing a couple of safety conditions: first, the allocation regions
of objects referred by the instance variables should outlive the
allocation region of the instance itself, and second, the allocation
regions of a class type and its superclass type must be the same.

@@

Note that if the original program in $\absof{\FB}$ contains unsafe
references, for example, a reference from a transferable region to a
stack regions, then the constraints generated during the elaboration
are not satisfiable. In such case, $\solve$ fails to solve
constraints, causing the program to be rejected.

@@

% Validity constraints primarily result from checking if a region type
% template of form $B\inang{\tbar}\inang{\rhoalloc,\rhobar}$ is
% well-formed (i.e., $\typeOk$). The corresponding well-formedness rule
% (Fig.~\ref{fig:fb-staticsem}) requires that the instantiation of $B$'s
% formal region parameters with actual region arguments satisfy its
% precondition, where actuals are subsituted for formals. Since $B$'s
% precondition could be unknown, the requirement results in a in a
% validity constraint with predicate variables and pending
% substitutions. Other sources of validity constraints are calls to
% region-polymorphic methods, where similar situation arises. 
In general, validity constraints generated by our algorithm assume one
of the following two forms:
\begin{center}
\( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
           {\phicstr} \qquad
   \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
           {F_j(\varphi_j)}
\)
\end{center}
Where $\varphi_i$'s denote the unknown preconditions of the class and
the method under elaboration. If the constraint is generated while
checking the well-formedness of a type or elaborating an expression,
then $\varphi_j$'s denote the unknown preconditions of classes and
methods that were used in that type or expression.  Each use of a
(region-polymorphic) class or a method may instantiate region
parameters differently, resulting in a different pending substitution
($F_j$). Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
of predicate variables and pending substitutions. While $\phictxt$
captures relationships that are \emph{known} to hold between concrete
region identifiers (i.e., $\rgn$'s) when the constraint was generated,
$\phicstr$ captures relationships that are \emph{required} to hold
among region varibles (i.e., $\rho$'s), or relationships between
region variables and identifiers. Each region variable occuring in
$\phicstr$ has an associated well-formedness constraint, which
specifies its unification domain. The unification domain of a
constraint is the union of unification domains of all region variables
occuring in the constraint.

>>>>

Region type templates are FGJ types extended with fresh region
variables ($\rho$) and predicate variables ($\varphi$) to denote
unknown region annotations and region constraints, respectively. For
instance, if a variable $x$ has type $\ObjZ$ in FGJ, its region type
template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
region variable. Likewise, given the region-annotated definition of
\C{Pair} class from \S\ref{sec:fb-syntax} a region type template for a
method with FGJ type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$
is \footnote{In our exposition, we assume that classes $\C{A}$ and
$\C{B}$ are trivial subclasses of $\ObjZ$ with no fields/methods. Like
$\ObjZ$, they accept one region parameter - the allocation region of
their objects.}\footnote{We abuse arrow notation to also represent
types of methods, but unlike function types, there is no allocation
region annotation atop the arrow in a method type.}
$\inang{\rhoalloc_0,\rho_{1-4} \,|\, \varphi_0}
\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_{1-3}} \rightarrow
\C{A}\inang{\rho_4}$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
variables, and $\varphi_0$ is a fresh predicate variable denoting
unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.  Region type
template of a type variable $\C{a}$ is $\C{a}@\rho$, where $\rho$ is
fresh. For a class, a template is computed in two steps. In the first
step, we templatize the types of all its fields, constructor
arguments, and arguments and return values of all its methods, along
with the type of its superclass. In the second step, we generalize all
the free region variables occuring in the templates computed in the
first step as region parameters of the class. Finally, we add a fresh
allocation region parameter ($\rhoalloc$) to the list of parameters,
and introduce a new predicate variable ($\varphi$) to denote unknown
constraints on region parameters. For example, consider the standard
FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
respectively. It can be templatized as following (method definitions
are left intact while templatizing, hence they are elided below):
\vspace*{-0.1in}
\begin{center}
\begin{codejava}
class Pair$\inang{a\extends\ObjZ, b\extends\ObjZ}$
          $\inang{\rhoalloc_0,\rho_{0-4}\,|\,\varphi_0} \extends \ObjZ\inang{\rho_4}$ {
  $a@\rho_0$ fst; $b@\rho_1$ snd;
  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { $...$ }
  a@$\rho_5$ getFst$\inang{\rhoalloc_1,\rho_5}$() { $...$ }
}
\end{codejava}
\end{center}
\vspace*{-0.1in}
% \footnote{In reality, the generalized region variables only come
% from the type templates of class's fields and superclass. Method
% types do not contain any free region variables.}
% The template elides $\C{Pair}$'s methods, whose region type
% templates contain no free variables. Among the region parameters of
% the class template, $\rho_{0-4}$ are obtained by generalizing free
% region variables in the types of its class's fields, constructor
% arguments, and its superclass type.  The remaining parameter
% ($\rhoalloc_0$) is a fresh region variable denoting the allocation
% region argument. Fresh predicate variable $\varphi_0$ denotes
% unknown constraints over $\rhoalloc_0$ and $\rho_{0-4}$ that need to
% hold for template to be a well-formed region-annotated class
% definition in $\FB$.

The type template for a recursively defined class is computed in two
steps. First, all recursive occurances of the class among the types of
its fields are ignored and the class is templatized as if it is a
non-recursive class. Next, all the recursive occurances are
templatized with respect to the class template computed in the first
step, such that their region annotations are \emph{exactly} same as
the region parameters of the class. Applying this technique on a
recursively defined $\C{LinkedListNode}\inang{\C{a}}$ class, for
example, results in a template that uses same region variable to
denote allocation regions of the current node, and the node that the
current node refers to. 
% For example, consider a generic $\C{ListNode}\inang{\C{a}}$ class
% containing two fields: $\C{data}$ of type $\C{a}$ and $\C{next}$ of
% type $\C{ListNode}\inang{\C{a}}$.  where $\C{a}$ is the type of the
% data stored in the linked list node.  To templatize the
% $\C{ListNode}\inang{a}$ class, we first ignore its recursive
% occurance in the type of $\C{next}$ field, and templatize the type
% $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this type
% template of $\C{data}$ field, we compute the class's template as
% following: \begin{center} $\C{ListNode}\inang{\C{a}\extends\ObjZ}
% \inang{\rhoalloc_0,\rho_0\,|\,\varphi_0} \extends \ObjZ \{...\}$
% \end{center} Next, we trivially templatize the type of $\C{next}$
% field in the body of the class as
% $\C{ListNode}\inang{\C{a}}\inang{\rhoalloc_0,\rho_0}$.  The
% resulting class represents a linked list with spine in the region
% $\rhoalloc_0$ and data objects in the region $\rho_0$.
% Observe that our strategy of trivially templatizing the recursive
% occurances of a class type in its definition ensures that no new
% region variables are introduced.
This technique can be extended to mutually recursive definitions in a
straightforward manner, by simultaneously templatizing them. Using the
techniques outlined above, we compute region type templates for all
classes bound in the class table of the $\absof{\FB}$ program before
we proceed to elaborate expressions.

@@

Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
this elaboration for a subset of expressions in $\absof{\FB}$
%, whose corresponding $\FB$ expressions have been ascribed static
%semantics in Fig.~\ref{fig:fb-staticsem}. 
$\elabExpr$ is defined under the same context as the expression typing
judgment in Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$,
and $\env$ retaining their meaning. The function traverses expressions
in a syntax-directed manner of a type checker, introducing fresh
region type templates for unknown region types, while generating
constraints over region and predicate variables. The precise nature of
generated constraints is explained in \S\ref{sec:fb-constraintsem},
but in summary, they capture the relationships between the type
templates of various subexpressions and their well-formedness. Note
that $\elabExpr$ returns the region type template of the
subexpression, which is used to generate constraints for the
expression. Functions $\typeOk$ and $\subtypeOk$ (definitions not
shown) used by $\elabExpr$ implement type well-formedness and subtype
judgments from Fig.~\ref{fig:fb-staticsem}, respectively.

@@

Functions $\elabMeth$ and $\elabClass$ shown in
Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
class definitions, respectively. Both functions first build a context
($\A$) containing a set ($ \rhoenv$) of regions that are currently
live, a map ($\aenv$) mapping type variables to their bounds, and a
constraint formula ($\phicx$) capturing constraints over live region
variables. We use predicate variables ($\varphi$ and $\varphi_m$) to
capture constraints over variables in $\rhoenv$ denoting the fact that
such constraints are yet to be inferred. Both elaboration functions
generate new constraints by checking well-formedness and subtype
relationships among types of their constituents. A fresh constraint
($C_4$) generated by $\elabClass$ captures a couple of safety
conditions: first, the allocation regions of objects referred by the
instance variables should outlive the allocation region of the
instance itself, and second, the allocation regions of a class type
and its superclass type must be the same.

% Function $\elabMeth$ elaborates a method definition of class $B$. It
% calls $\elabExpr$ with the context $\A$, its allocation context
% parameter ($\rhoallocm$), and a type environment ($\env$) that
% contains region type bindings for all the arguments of the method,
% including the implicit $\C{this}$ argument. The region type template
% returned by $\elabExpr$ for the method body is checked against its
% expected type (derived from the type template of the method)
% generating more constraints. The function then returns the elaborated
% method definition and the set of constraints.

% $\elabClass$ elaborates the definition of a class $B$. It relies on
% $\elabCons$ (not shown) and $\elabMeth$ functions to elaborate $B$'s
% constructor ($k$) and method definitions ($\bar{d}$), respectively. To
% the set of constraints returned by these functions, $\elabClass$ adds
% constraints generated by checking the well-formedness of the type
% templates of its superclass and fields, and also a new constraint
% capturing a couple of safety conditions: first, the allocation regions
% of objects referred by the instance variables should outlive the
% allocation region of the instance itself, and second, the allocation
% regions of a class type and its superclass type must be the same.

@@

% Note that if the original program in $\absof{\FB}$ contains unsafe
% references, for example, a reference from a transferable region to a
% stack regions, then the constraints generated during the elaboration
% are not satisfiable. In such case, $\solve$ fails to solve
% constraints, causing the program to be rejected.

@@

% Validity constraints primarily result from checking if a region type
% template of form $B\inang{\tbar}\inang{\rhoalloc,\rhobar}$ is
% well-formed (i.e., $\typeOk$). The corresponding well-formedness rule
% (Fig.~\ref{fig:fb-staticsem}) requires that the instantiation of $B$'s
% formal region parameters with actual region arguments satisfy its
% precondition, where actuals are subsituted for formals. Since $B$'s
% precondition could be unknown, the requirement results in a in a
% validity constraint with predicate variables and pending
% substitutions. Other sources of validity constraints are calls to
% region-polymorphic methods, where similar situation arises. 
In general, validity constraints generated by our algorithm assume one
of the following two forms:
\begin{center}
\( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
           {\phicstr} \qquad
   \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
           {F_j(\varphi_j)}
\)
\end{center}
Where $\varphi_i$'s denote the unknown preconditions of the class and
the method under elaboration. If the constraint is generated while
checking the well-formedness of a type or elaborating an expression,
then $\varphi_j$'s denote the unknown preconditions of classes and
methods that were used in that type or expression. Each use of a
(region-polymorphic) class or a method has to instantiate region
parameters, resulting in pending substitutions ($F_j$). Formulas
$\phictxt$ and $\phicstr$ are concrete, i.e., free of predicate
variables and pending substitutions. While $\phictxt$ captures
relationships that are \emph{known} to hold between concrete region
identifiers (i.e., $\rgn$'s) when the constraint was generated,
$\phicstr$ captures relationships that are \emph{required} to hold
among region varibles (i.e., $\rho$'s), or relationships between
region variables and identifiers. Each region variable occuring in
$\phicstr$ has an associated well-formedness constraint, which
specifies its unification domain. The unification domain of a
constraint is the union of unification domains of all region variables
occuring in the constraint.
